Approximation as Differentiation

### 2025-07-09 main/approx-as-tangents.tex
Fix autodiff ref.
Add bounded-meet to 'everything' and reference paper theorem.
Correction to chain rule.

\subsection{Manifolds, Smooth Functions, and Automatic Differentiation}

@@

\begin{remark}[Chain Rule]
  \label{rem:chain-rule}
  A useful property of derivative maps is that they compose according
  to the chain rule. Suppose that $f : M \to N$ and $g : N \to K$ are
  smooth functions. Then for any $x \in M$, we have:
  \begin{itemize}
  \item $\pushf{(g \circ f)}_x = \pushf{g}_{f(x)} \circ \pushf{f}_x : \tangents_x(M) \linearto \tangents_{g(f(x)}(K)$
  \item $\pullf{(g \circ f)}_x = \pullf{f}_x \circ \pullf{g}_{f(x)} : \cotangents_{f(x)}(K) \linearto \cotangents_x(M)$
  \end{itemize}
  The chain rule has the practical effect that we can compute
  derivative maps of $f$ and $g$ independently and compose them,
  instead of the potentially more difficult task of computing the
  derivative maps of $g \circ f$. As we shall see below, stable maps
  also obey a chain rule, and this forms the basis of the general
  categorical approach to differentiability that we describe in
  \secref{models-of-total-gps}.
\end{remark}

@@

\begin{theorem}
  Bounded meet lattices and conditionally multiplicatve functions form
  a category $\CM$. This category has products, coproducts, and
  exponentials.
\end{theorem}

@@

\begin{example}[Conditionally Multiplicative Functions]
  \label{ex:strict-short-circuit}
  To see the effect of conditional multiplicativity, consider several
  ways of defining the OR on the lifted booleans
  $\mathbb{B}_\bot$. Two functions that are cm are the strict and
  left-short-circuiting ORs\footnote{The clauses in these examples are
    shorthand for the graph of the function. They are not to be
    understood as pattern matching clauses in a language like Haskell,
    where it is not possible to match on $\bot$.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
    \qquad
    \raisebox{-15ex}{
      \begin{tikzpicture}
        \node (top) at (0,0) {$(\mathsf{tt},\mathsf{ff})$};
        \node [below of=top,xshift=-0.5cm] (oi) {$(\bot,\mathsf{ff})$};
        \node [below of=top,xshift=0.5cm] (io) {$(\mathsf{tt},\bot)$};
        \node [below of=oi,xshift=0.5cm] (oo) {$(\bot,\bot)$};
        \draw (top) -- (oi);
        \draw (top) -- (io);
        \draw (oi) -- (oo);
        \draw (io) -- (oo);
      \end{tikzpicture}
    }
  \end{displaymath}
  In the poset $\mathbb{B}_\bot^2$, a typical poset of approximations of a fully defined element is shown to the right. For $\mathrm{strictOr}$, any approximation that isn't the fully defined input is mapped to $\bot$, while $\mathrm{shortCircuitOr}$ maps the partially defined $(\mathsf{tt},\bot)$ to $\mathsf{tt}$. Thus, even though these functions operate identically on fully defined inputs, they differ in their \emph{derivatives} on partially defined input, exposing how . That they are cm can be checked by examining their restrictions' behaviour. If we take the approximations $(\bot,\mathsf{ff})$ and $(\mathsf{tt},\bot)$, then their meet is $(\bot,\bot)$; $\mathrm{strictOr}$ maps all three elements to $\bot$, so is cm here since $\bot \wedge \bot = \bot$; and $\mathrm{shortCircuitOR}$ has $(\bot,\mathsf{ff}) \mapsto \bot$ and $(\mathsf{tt},\bot) \mapsto \mathsf{tt}$, the meet of which is $\bot = \mathrm{shortCircuitOR}(\bot,\bot)$. Other combinations can be checked similarly.
\end{example}

>>>>

\subsection{Manifolds, Smooth Functions, and Automatic Differentiation}
\label{sec:approx-as-tangents:autodiff}

@@

\begin{remark}[Chain Rule]
  \label{rem:chain-rule}
  A useful property of derivative maps is that they compose according
  to the chain rule. Suppose that $f : M \to N$ and $g : N \to K$ are
  smooth functions. Then for any $x \in M$, we have:
  \begin{itemize}
  \item $\pushf{(g \circ f)}_x = \pushf{g}_{f(x)} \circ \pushf{f}_x : \tangents_x(M) \linearto \tangents_{g(f(x))}(K)$
  \item $\pullf{(g \circ f)}_x = \pullf{f}_x \circ \pullf{g}_{f(x)} : \cotangents_{g(f(x))}(K) \linearto \cotangents_x(M)$
  \end{itemize}
  The chain rule has the practical effect that we can compute
  derivative maps of $f$ and $g$ independently and compose them,
  instead of the potentially more difficult task of computing the
  derivative maps of $g \circ f$. As we shall see below, stable maps
  also obey a chain rule, and this forms the basis of the general
  categorical approach to differentiability that we describe in
  \secref{models-of-total-gps}.
\end{remark}

@@

\begin{theorem}
  Bounded meet lattices and conditionally multiplicative functions form
  a category $\CM$. This category has products, coproducts, and
  exponentials.
\end{theorem}

@@

\begin{example}[Conditionally Multiplicative Functions]
  \label{ex:strict-short-circuit}
  To see the effect of conditional multiplicativity, consider several
  ways of defining the OR on the lifted booleans
  $\mathbb{B}_\bot$. Two functions that are cm are the strict and
  left-short-circuiting ORs\footnote{The clauses in these examples are
    shorthand for the graph of the function. They are not to be
    understood as pattern matching clauses in a language like Haskell,
    where it is not possible to match on $\bot$.}:
  \begin{displaymath}
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{tt}&\mathsf{ff}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{tt}&=&\mathsf{tt} \\
      \mathrm{strictOr}&\mathsf{ff}&\mathsf{ff}&=&\mathsf{ff} \\
      \mathrm{strictOr}&\bot&\_&=&\bot \\
      \mathrm{strictOr}&\_&\bot&=&\bot \\
    \end{array}
    \qquad
    \begin{array}[t]{l@{(}l@{,~}l@{)~}c@{~}l}
      \mathrm{shortCircuitOR}&\mathsf{tt}&\_&=&\mathsf{tt} \\
      \mathsf{shortCircuitOR}&\mathsf{ff}&x&=&x \\
      \mathsf{shortCircuitOR}&\bot&\_& =& \bot
    \end{array}
    \qquad
    \raisebox{-15ex}{
      \begin{tikzpicture}
        \node (top) at (0,0) {$(\mathsf{tt},\mathsf{ff})$};
        \node [below of=top,xshift=-0.5cm] (oi) {$(\bot,\mathsf{ff})$};
        \node [below of=top,xshift=0.5cm] (io) {$(\mathsf{tt},\bot)$};
        \node [below of=oi,xshift=0.5cm] (oo) {$(\bot,\bot)$};
        \draw (top) -- (oi);
        \draw (top) -- (io);
        \draw (oi) -- (oo);
        \draw (io) -- (oo);
      \end{tikzpicture}
    }
  \end{displaymath}
  In the poset $\mathbb{B}_\bot^2$, a typical poset of approximations of a fully defined element is shown to
  the right. For $\mathrm{strictOr}$, any approximation that isn't the fully defined input is mapped to
  $\bot$, while $\mathrm{shortCircuitOr}$ maps the partially defined $(\mathsf{tt},\bot)$ to $\mathsf{tt}$.
  Thus, even though these functions operate identically on fully defined inputs, they differ in their
  \emph{derivatives} on partially defined input, exposing how they explore their arguments differently. That
  they are cm can be checked by examining their restrictions' behaviour. If we take the approximations
  $(\bot,\mathsf{ff})$ and $(\mathsf{tt},\bot)$, then their meet is $(\bot,\bot)$; $\mathrm{strictOr}$ maps
  all three elements to $\bot$, so is cm here since $\bot \wedge \bot = \bot$; and $\mathrm{shortCircuitOR}$
  has $(\bot,\mathsf{ff}) \mapsto \bot$ and $(\mathsf{tt},\bot) \mapsto \mathsf{tt}$, the meet of which is
  $\bot = \mathrm{shortCircuitOR}(\bot,\bot)$. Other combinations can be checked similarly.
\end{example}

### 2025-07-09 main/introduction.tex
Fix Vakar citation.
Correction to chain rule.

Existing approaches to program slicing are often tied to particular programming languages or implementations. In this paper we develop a general categorical approach to program slicing and \roly{complete}.

@@

\begin{example}
  \label{ex:introduction-example}
  The following program is written in Haskell \cite{haskell} syntax, using a list comprehension to filter a list of pairs of labels and numbers to those numbers with a given label, and then to compute the sum of the numbers:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{query} :: \mathrm{Label} \to [(\mathrm{Label}, \mathrm{Int})] \to \mathrm{Int} \\
      \mathrm{query}\,l\,\mathit{db} = \mathrm{sum}\,[ n \mid (l',n) \leftarrow \mathit{db}, l \equiv l' ]
    \end{array}
  \end{displaymath}
  With $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, we will have $\mathit{query}\,\mathsf{a}\,\mathit{db}$ and $\mathit{query}\,\mathsf{b}\,\mathit{db}$ both evaluating to $1$.

@@

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (ooi);
      \draw (oii) -- (oio);
      \draw (iio) -- (ioo);
      \draw (iio) -- (oio);
      \draw (ooi) -- (bot);
      \draw (oio) -- (bot);
      \draw (ioo) -- (bot);
    \end{tikzpicture}
  \end{center}
  The output approximation lattice looks like this, where $1$ is the actual data point that was returned in both runs of the program, and $\bot$ indicates that we are approximating this piece of data away:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1$};
      \node [below of=top] (bot) {$\bot$};
      \draw (top) -- (bot);
    \end{tikzpicture}
  \end{center}
  These are not the only choices of approximation lattices that we could have taken. For the input, we have chosen a lattice that allows us to ``forget'' (approximate away) numbers in the input, but not any of the other data or the overall structure. However, other choices are also useful. Indeed, one of the aims of this work is to clarify how to choose an approximation structure appropriate for different tasks by use of type information. We elaborate on this further in \secref{first-order}.

@@

  In this case, the forwards portion of the Galois connection tells us for each approximation of the input whether or not it is sufficient to compute the output. In this example, if we provide insufficient data to compute the output, then we will get an underapproximated output. For example, $\partial (\mathit{query}\, \mathsf{a})_f([(\mathsf{a},0),(\mathsf{b},\bot),(\mathsf{a},\bot)]) = \bot$ because we need all the values associated with the label $\mathsf{a}$ to compute their sum.

>>>>

Existing approaches to program slicing are often tied to particular programming languages or implementations. In this paper we develop a general categorical approach to program slicing and \todo{complete}.

@@

\begin{example}
  \label{ex:introduction-example}
  The following program is written in Haskell \cite{haskell} syntax, using a list comprehension to filter a list of pairs of labels and numbers to those numbers with a given label, and then computing the sum of the numbers:
  \begin{displaymath}
    \begin{array}{l}
      \mathrm{query} :: \mathrm{Label} \to [(\mathrm{Label}, \mathrm{Int})] \to \mathrm{Int} \\
      \mathrm{query}\,l\,\mathit{db} = \mathrm{sum}\,[ n \mid (l',n) \leftarrow \mathit{db}, l \equiv l' ]
    \end{array}
  \end{displaymath}
  With $\mathit{db} = [(\mathsf{a}, 0), (\mathsf{b}, 1), (\mathsf{a}, 1)]$, we will have $\mathit{query}\,\mathsf{a}\,\mathit{db}$ and $\mathit{query}\,\mathsf{b}\,\mathit{db}$ both evaluating to $1$.

@@

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (ooi);
      \draw (oii) -- (oio);
      \draw (iio) -- (ioo);
      \draw (iio) -- (oio);
      \draw (ooi) -- (bot);
      \draw (oio) -- (bot);
      \draw (ioo) -- (bot);
    \end{tikzpicture}
  \end{center}
  In both runs of the program, the output approximation lattice looks like this, where $1$ is the actual data point that was returned, and $\bot$ indicates that we are approximating this piece of data away:
  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1$};
      \node [below of=top] (bot) {$\bot$};
      \draw (top) -- (bot);
    \end{tikzpicture}
  \end{center}
  These are not the only choices of approximation lattices that we could have made. For the input, we have chosen a lattice that allows us to ``forget'' (approximate away) numbers in the input, but not any of the other data or the overall structure. However, other choices are also useful. Indeed, one of the aims of this work is to clarify how to choose an approximation structure appropriate for different tasks by use of type information. We elaborate on this further in \secref{models-of-total-gps}.

@@

  In this case, the forwards portion of the Galois connection tells us for each approximation of the input whether or not it is sufficient to compute the output. If we provide insufficient data to compute the output, then we will get an underapproximated output. For example, $\partial (\mathit{query}\, \mathsf{a})_f([(\mathsf{a},0),(\mathsf{b},\bot),(\mathsf{a},\bot)]) = \bot$ because we need all the values associated with the label $\mathsf{a}$ to compute their sum.
